\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}

### 2025-05-07 misc/JSS article/final/article.tex
Update benchmarks.

1   collapse   3.61µs   4.26µs   14.33KB  9999    1    43.81ms
2      dplyr 211.27µs 240.26µs    3.92MB  7651   52      1.86s

@@

\pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/summary-statistics.html}{summary statistics functions} offer a parsimonious yet powerful toolkit to examine complex datasets. A particular focus has been on providing tools for examining longitudinal (panel) data. Recall the indexed World Development panel dataset (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}). The function \fct{varying} can be used to examine which of these variables are time-varying.

@@

A related exercise is to decompose the variance of a panel series into variation between countries and variation within countries over time. Using the (de-)meaning functions \fct{fbetween}/ \fct{fwithin} supporting \class{indexed\_series} (see Table~\ref{tab:TSfun}), this is easily demonstrated.

@@

R> LIFEEXi <- reindex(wlddev$LIFEEX, wlddev$iso3c)

@@

%
 It also has a \code{stepwise} argument to describe one variable at a time, allowing users to naturally 'click-through' the variables in a large dataset while spreading the computational burden. The \href{https://sebkrantz.github.io/collapse/reference/descr.html}{documentation} provides more details and examples. Both \fct{qsu} and \fct{descr} come with an \code{as.data.frame()} method for efficient tidying and easy further analysis. \newline

@@

1     base_int 41.49ms  43.2ms   166.2MB    47   47      2.04s
2 collapse_int  5.47ms  9.28ms    38.2MB   220   44      2.01s

@@

1     base_char 68.7ms   70ms   166.2MB    28   28      2.01s
2 collapse_char 13.8ms 15.7ms    38.2MB   123   21      2.01s

@@

1     base_int 15.13ms 15.58ms    76.3MB    38   67   608.03ms
2 collapse_int  5.47ms  7.74ms    38.2MB   196   49      1.43s

@@

1       base_char 60.8ms 62.4ms   114.5MB    32   32      2.07s
2 data.table_char 31.9ms   34ms    38.1MB    58   18      2.02s
3   collapse_char 19.6ms 21.8ms    38.1MB    90   18      2.01s

@@

1          R   5.67ms   6.21ms    7.86KB   324    0         2s
2   collapse 280.93µs 321.64µs    7.86KB  6143    1         2s

@@

1          R   6.01ms   6.21ms   32.09KB   324    0         2s
2   collapse 278.92µs 322.14µs    7.86KB  6164    0         2s

@@

1         MS 78.7ms 79.7ms   238.9KB    26    0      2.07s
2   collapse 19.4ms 19.6ms    27.2KB   101    0      2.02s

@@

1          R   5.36ms 5.59ms    7.87MB   319   20      1.79s
2   collapse 923.24µs 1.14ms    7.71MB  1279   60      1.58s

@@

1      dplyr 437.44ms 471.82ms    50.5MB     5   50      2.35s
2 data.table   7.56ms   8.59ms    20.8MB   208   18         2s
3   collapse      3ms   3.75ms     9.2MB   467   21         2s

@@

1      dplyr_mean  1.17s  1.18s   48.57MB     2   42      2.36s
2 data.table_mean 7.64ms 8.79ms   18.93MB   190   18         2s
3   collapse_mean 3.28ms 3.77ms    9.11MB   470   17      2.01s

@@

1      dplyr_median   4.05s   4.05s    52.8MB     1   86      4.05s
2 data.table_median 21.45ms 22.51ms    18.9MB    86    8      2.01s
3   collapse_median   9.7ms 10.34ms    11.1MB   171   11         2s

@@

1      dplyr 90.25ms 99.01ms   20.18MB    20   44      2.09s
2 data.table 59.54ms 63.14ms    5.77MB    32   27      2.04s
3   collapse  5.32ms  5.88ms     6.8MB   303   11         2s

@@

1      dplyr_joins 140.85ms 173.7ms   559.4MB    12   28      2.12s
2 data.table_joins  99.45ms 132.9ms     491MB    15   36      2.11s
3   collapse_joins   9.15ms  11.2ms    89.7MB   142   40      2.01s

@@

Last but not least, I benchmark pivots, starting with a long pivot that simply melts the 6 columns aggregated beforehand into one column, duplicating all other columns 6 times:
%
\begin{Schunk}
\begin{Sinput}
R> bmark(tidyr = tidyr::pivot_longer(flights, cols = vars),
+        data.table = qDT(flights) |> melt(measure = vars),
+        collapse = pivot(flights, values = vars))
\end{Sinput}
\begin{Soutput}
  expression    min median mem_alloc n_itr n_gc total_time
1      tidyr 58.2ms 58.2ms     255MB     1   27     58.2ms
2 data.table 41.9ms 43.9ms     209MB    14   22    613.6ms
3   collapse   13ms 15.3ms     209MB    34   57    512.9ms
\end{Soutput}
\end{Schunk}
%
Memory-wise, \pkg{collapse} and \pkg{data.table} are equally efficient, but \pkg{collapse} is faster, presumably due to more extensive use of \code{memset()} to copy values in \proglang{C} or smaller \proglang{R}-level overheads.

To complete the picture, I also also perform a wide pivot where the 6 columns are summed (for efficiency) across the 3 origin airports and expanded to create 18 airport-value columns.
%
\begin{Schunk}
\begin{Sinput}
R> bmark(tidyr = tidyr::pivot_wider(flights, id_cols = .c(month, day, dest),
+            names_from = "origin", values_from = vars, values_fn = sum),
+        data.table = dcast(qDT(flights), month + day + dest ~ origin,
+                           value.var = vars, fun = sum),
+        collapse_fsum = pivot(flights, .c(month, day, dest), vars,
+                              "origin", how = "wider", FUN = fsum),
+        collapse_itnl = pivot(flights, .c(month, day, dest), vars,
+                              "origin", how = "wider", FUN = "sum"))
\end{Sinput}
\begin{Soutput}
     expression      min   median mem_alloc n_itr n_gc total_time
1         tidyr 353.82ms 383.22ms     143MB     5   37         2s
2    data.table 222.83ms 230.79ms    21.7MB     9   40      2.08s
3 collapse_fsum   5.04ms   7.01ms    39.1MB   233   23      2.07s
4 collapse_itnl   4.12ms   4.51ms    12.4MB   392    8      2.01s
\end{Soutput}
\end{Schunk}
%
Again, \pkg{collapse} is fastest, as it offers full vectorization, either via \fct{fsum}, which translates to \code{fsum(x, g, TRA = "fill")} before pivoting and thus entails a full deep copy of the \code{vars} columns, or via an optimized internal sum function which sums values 'on the fly' during the reshaping process. \pkg{data.table} is not vectorized here but at least memory efficient. %\newline

>>>>

1   collapse   3.32µs   4.18µs   14.33KB  9999    1    43.08ms
2      dplyr    200µs 238.91µs    1.53MB  9934   66      2.48s

@@

\pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/summary-statistics.html}{summary statistics functions} offer a parsimonious yet powerful toolkit to examine complex datasets. A particular focus has been on providing tools for examining longitudinal (panel) data. Recall the World Development panel dataset (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) from Section~\ref{sec:integration}. The function \fct{varying} can be used to examine which of these variables are time-varying.

@@

A related exercise is to decompose the variance of a panel series into variation between countries and variation within countries over time. Using the (de-)meaning functions \fct{fbetween}/ \fct{fwithin} supporting \class{indexed\_series} (see Table~\ref{tab:TSfun}), this is easily demonstrated:

@@

R> LIFEEXi <- reindex(wlddev$LIFEEX, index = wlddev$iso3c)

@@

%
 It also has a \code{stepwise} argument to describe one variable at a time, allowing users to naturally 'step-through' the variables in a large dataset while spreading the computational burden. The \href{https://sebkrantz.github.io/collapse/reference/descr.html}{documentation} provides more details and examples. Both \fct{qsu} and \fct{descr} come with an \code{as.data.frame()} method for efficient tidying and easy further analysis. \newline

@@

1     base_int 43.78ms  45.3ms   166.2MB    65   65         3s
2 collapse_int  6.25ms  8.24ms    38.2MB   353   71         3s

@@

1     base_char 69.9ms 71.3ms   166.2MB    43   43      3.07s
2 collapse_char 13.9ms 15.6ms    38.2MB   180   36      3.01s

@@

1     base_int 16.92ms  17.4ms    76.3MB    78   78      1.37s
2 collapse_int  6.95ms  7.68ms    38.2MB   270   75      2.08s

@@

1       base_char 58.8ms 58.8ms   114.5MB     1   48    58.78ms
2 data.table_char 32.2ms 33.7ms    38.1MB    66   19      2.22s
3   collapse_char 19.5ms 20.4ms    38.1MB   113   28      2.31s

@@

1          R   5.4ms   6.2ms    7.86KB   493    0         3s
2   collapse 282.9µs 324.5µs    7.86KB  9150    1         3s

@@

1          R   5.38ms   6.2ms   32.09KB   491    0         3s
2   collapse 280.93µs 327.2µs    7.86KB  8975    0         3s

@@

1         MS 77.8ms 79.4ms    89.9KB    38    0      3.02s
2   collapse 19.1ms 19.5ms    27.2KB   154    0      3.01s

@@

1          R   5.4ms 5.62ms    7.87MB   459   35      2.61s
2   collapse 900.2µs 1.13ms    7.71MB  1876  104      2.19s

@@

1      dplyr 363.83ms 479.56ms    50.5MB     7   65      3.15s
2 data.table   7.22ms   7.99ms    20.8MB   339   33         3s
3   collapse   2.96ms    3.4ms     9.2MB   769   33         3s

@@

1      dplyr_mean  1.16s  1.17s   48.57MB     3   78      3.52s
2 data.table_mean 7.43ms 8.31ms   18.93MB   283   28         3s
3   collapse_mean  3.1ms 3.71ms    9.11MB   679   28         3s

@@

1      dplyr_median    4.2s    4.2s    52.8MB     1   91       4.2s
2 data.table_median 21.32ms 22.24ms    18.9MB   130   13         3s
3   collapse_median  9.34ms  9.96ms    11.1MB   271   12         3s

@@

1      dplyr    93ms 103.73ms   20.18MB    27   56      3.03s
2 data.table 56.51ms  60.06ms    5.77MB    50   25      3.06s
3   collapse  5.04ms   5.51ms     6.8MB   509   13         3s

@@

1      dplyr_joins 139.95ms   153ms   559.4MB    19   32      3.12s
2 data.table_joins  97.39ms   129ms     491MB    19   39      3.11s
3   collapse_joins   8.85ms  11.9ms    89.7MB   188   51      3.01s

@@

Last but not least, I benchmark pivots, starting with a long pivot that simply melts the 6 columns aggregated beforehand into one column, duplicating all other columns 6 times:
%
\begin{Schunk}
\begin{Sinput}
R> bmark(tidyr = tidyr::pivot_longer(flights, cols = vars),
+        data.table = qDT(flights) |> melt(measure = vars),
+        collapse = pivot(flights, values = vars))
\end{Sinput}
\begin{Soutput}
  expression    min median mem_alloc n_itr n_gc total_time
1      tidyr 56.9ms 58.8ms     254MB     8   39      472ms
2 data.table   40ms 42.6ms     209MB    22   32      962ms
3   collapse   13ms   15ms     209MB    47   84      713ms
\end{Soutput}
\end{Schunk}
%
Memory-wise, \pkg{collapse} and \pkg{data.table} are equally efficient, but \pkg{collapse} is faster, presumably due to more extensive use of \code{memset()} to copy values in \proglang{C} or smaller \proglang{R}-level overheads.

To complete the picture, I also also perform a wide pivot where the 6 columns are summed (for efficiency) across the 3 origin airports and expanded to create 18 airport-value columns.
%
\begin{Schunk}
\begin{Sinput}
R> bmark(tidyr = tidyr::pivot_wider(flights, id_cols = .c(month, day, dest),
+            names_from = "origin", values_from = vars, values_fn = sum),
+        data.table = dcast(qDT(flights), month + day + dest ~ origin,
+                           value.var = vars, fun = sum),
+        collapse_fsum = pivot(flights, .c(month, day, dest), vars,
+                              "origin", how = "wider", FUN = fsum),
+        collapse_itnl = pivot(flights, .c(month, day, dest), vars,
+                              "origin", how = "wider", FUN = "sum"))
\end{Sinput}
\begin{Soutput}
     expression      min   median mem_alloc n_itr n_gc total_time
1         tidyr 379.96ms 397.91ms     143MB     8   59      3.27s
2    data.table    227ms  231.1ms    21.7MB    13   47      3.03s
3 collapse_fsum   5.61ms   7.18ms    39.1MB   327   26      3.01s
4 collapse_itnl   3.77ms   4.45ms    12.4MB   611   15         3s
\end{Soutput}
\end{Schunk}
%
Again, \pkg{collapse} is fastest, as it offers full vectorization, either via \fct{fsum}, which translates to \code{fsum(x, g, TRA = "fill")} before pivoting and thus entails a full deep copy of the \code{vars} columns, or via an optimized internal sum function which sums values 'on the fly' during the reshaping process. \pkg{data.table} is not vectorized here but at least memory efficient. %\newline
